---
title: "A Brief Guide to Stan's Warnings"
author: ""
date: "July 26, 2016"
output: 
  html_document:
    toc: true
    toc_depth: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Stan emits a lot of warning messages. In this document we go over why this
happens, explain some of the most common warnings, and provide some tips about
how to proceed depending on the particular warnings you're seeing.


### Why does Stan give so many warnings?

Stan throws a lot of warnings because the warnings are important. This 
does not mean that _every_ time you see a warning you can't trust your estimates, 
but when you see warnings it does mean that you shouldn't blindly trust 
your estimates without first understanding what the warnings mean and which 
ones require action on your part. 

_This is true of all software used to estimate complex statistical models_. You 
might not be used to seeing so many warnings from other software you use, but 
that _does not mean_ that Stan has more problems than that other software. The 
Stan Development Team places a high priority on notifying users about any issue 
that could potentially be important and Stan will always tell you about problems
it encounters instead of hiding them from you. Warning messages do not indicate 
that something is wrong with Stan but rather that Stan is doing its job and 
warning you when it finds problems (or possible problems) with what it was told
to do.



### Common warnings

#### Divergent transitions after warmup

**Example:**

```r
There were 15 divergent transitions after warmup. Increasing adapt_delta above 0.8 may help. 
```

Hamiltonian Monte Carlo (HMC) works by simulating the evolution of a [Hamiltonian 
system](https://en.wikipedia.org/wiki/Hamiltonian_system). Intuitively, 
you can think of simulating the movement of a partical, in our case a partical
that is exploring the high-dimensional posterior distribution defined by 
a Stan program + data. 

However, in order to approximate the exact solution of the Hamiltonian dynamics 
we need to choose 
a step size governing how far we move each time we evolve the system forward. 
When the step size is too large relative to the curvature of the log posterior 
this approximation can _diverge_ and threaten the validity of the sampler.

For some intuition, imagine walking down a steep mountain. If you take too
big of a step you will fall, but if you can take very tiny steps you might 
be able to make your way down the mountain, albeit very slowly. Similarly, 
we can tell Stan to take smaller steps around the posterior distribution, 
which (in some but not all cases) can help avoid these divergences. 
  
If the number of divergences is small then the first thing we recommend trying 
is increasing the value of the `adapt_delta` parameter (see below for example
code). This is the target average proposal acceptance probability during Stan's
adaptation period, and increasing it will force Stan to take smaller steps. The
downside is that sampling will tend to be slower because a smaller step size
means that more steps are required. **Since the validity of the estimates is not
guaranteed if there are post-warmup divergences, the slower sampling is a minor
cost.**

Unfortunately, for some models simply lowering the step size is not sufficient. 
Sometimes the geometry of the posterior distribution is such that only by 
reparameterizing the model will we be able to get valid estimates. That is, we
have to find a different way to write the model that is equivalent statistically
but that simplifies the geometry of the posterior distribution. This problem 
occurs frequently with hierarchical models and one of the simplest examples is 
Neal's Funnel, which is discussed in the *Optimizing Stan Code* chapter of the
[Stan manual](http://mc-stan.org/documentation/).

**Recommendations:**

* Increase the target acceptance rate `adapt_delta`. In RStan, `adapt_delta` is 
one of the parameters that you can include in the optional `control` list passed
to the `stan` or `sampling` functions. For example, to set `adapt_delta`
to 0.99 (the default is 0.8) you would do this:

```{r, eval=FALSE}
stan(..., control = list(adapt_delta = 0.99))
```

* Reparameterize your model. See the _Reparameterization_ section of the 
_Optimizing Stan Code_ chapter in the [Stan 
manual](http://mc-stan.org/documentation/), in particular the discussion of
hierarhical models and non-centered parameterizations.

<br>

#### Exception ... metropolis proposal rejected

**Example:**

```r
Exception thrown at line 24: normal_log: Scale parameter is 0, but must be positive! 
```

This warning indicates that the standard deviation (scale parameter) of the 
normal distribution (at line 24 in the Stan program) is 0, but it must be
positive for Stan to compute the value of density function. It is not uncommon
to get this type of warning if your model has bounded parameters. A message like
this does not necessarily indicate that something is wrong with your model, as
it's possible that even in a correctly specified and well-behaved model the
value of the scale parameter can be 0 (or numerically indistinguishable from 0)
occasionally during sampling.

However, this warning should _not_ be ignored if it occurs many times. 
Informally, if the number of times this happens is large then something in the
model or data (or the combination of the model and data) is consistently forcing
the constrained parameter to its boundary and this should be investigated.

**Recommendations:** 

* Ignore if very sporadic.
* Simulate data and fit the same model to check whether the same problem occurs
using different data.
* Use priors that place less weight on values near the boundary.


<br>

#### Maximum treedepth exceeded

Warnings about hitting the maximum treedepth are not as serious as warnings 
about divergent transitions. While divergent transitions are a _validity_
concern, hitting the maximum treedepth is an _efficiency_ concern. Configuring
the No-U-Turn-Sampler (the variant of HMC used by Stan) involves putting a cap
on the depth of the trees that it evaluates during each iteration (for details
on this see the *Hamiltonian Monte Carlo Sampling* chapter in the [Stan 
manual](http://mc-stan.org/documentation/)). This is controlled through a 
maximum depth parameter `max_treedepth`. When the maximum allowed tree depth is 
reached it indicates that NUTS is terminating prematurely to avoid excessively 
long execution time.

**Recommendations:** 

* Increase the maximum allowed treedepth. In RStan, `max_treedepth` is one
of the parameters that you can include in the optional `control` list passed
to the `stan` or `sampling` functions. For example, to set `max_treedepth`
to 15 (the default is 10) you would do this:

```{r, eval=FALSE}
stan(..., control = list(max_treedepth = 15))
```

